The Strucplot Framework:\\ Visualizing Multi-way Contingency Tables with \pkg{vcd}

### 2006-09-22 trunk/vcd/paper/Strucplot/JSS.Rnw
Change default of alternate_labeling in pairs_diagonal_mosaic
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@574 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

<<subsetting,results=verbatim>>=
(tit <- structable(~ Sex + Class + Age, data = Titanic))
tit["Male",]
tit["Male", c("1st","2nd","3rd")]
@

@@

Since mosaic displays are ``conditional plots'' by definition, we can
also use one single mosaic for stratified plots.
The formula interface of \codefun{mosaic} allows the
specification of conditioning variables (see Figure \ref{fig:conditioning}):

<<Conditioning>>=
mosaic(~ Age + Class | Sex, data = tit, split_vertical = TRUE, keep_aspect_ratio = FALSE)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Conditioningfig,fig=TRUE,echo=FALSE,width=12>>=
<<Conditioning>>
@
\caption{Mosaic plot again visualizing the 
distribution of class and age, given gender, this time using a
single mosaic plot. In contrast to Figure \ref{fig:parttable}, this
plot also visualizes the marginal distribution of `Sex'.}
\label{fig:conditioning}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7}

\noindent The effect of using this kind of formula is that
conditioning variables are permuted ahead of the the conditioned variables
in the table, and that \codefun{spacing\_conditional} is used as
default to better
distinguish conditioning from conditioned dimensions.
This spacing uses equal space between tiles of conditioned variables,
and increasing space between tiles of conditioning variables. In
addition, we release the fixed aspect ratio to get less distorted margins.

@@

<<cotabplot>>=
cotabplot(~ Age + Class | Sex, data = tit, panel_args = list(margins = 3), 
          labeling = labeling_left, tl_labels = c(Class = FALSE))
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=11>>=
<<cotabplot>>
@
\caption{Conditional table plot for the \data{Titanic} data,
  again visualizing the distribution of age and class, given
  gender, using separate mosaic displays like the `manual' plot 
  in Figure \ref{fig:parttable}.}
\label{fig:cotabplot}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

\noindent The \code{labeling\_args} argument modifies the labels'
appearance: here, to be left-aligned and unclipped
(see Section \ref{sec:labeling}).

Another high-level function for visualizing conditional independence
models are the \codefun{pairs} methods for table and structable objects.
In contrast to \codefun{cotabplot} which conditions on variables,
the \codefun{pairs} methods create pairwise views of the table. 
The function produces, by default, a plot matrix 
having strucplot displays in the off-diagonal panels, and
the variable names (optionally, with univariate statistics) in the diagonal cells. 
Figure \ref{fig:pairs} shows a pairs display for the
\data{HairEyeColor} data with univariate mosaics in the diagonal,
and mosaic plots visualizing mutual independence in the upper and lower triangles.
Due to the inherent asymmetry of mosaic displays, the corresponding plots in the
upper and lower triangle differ depending on which variable is used first
for splitting---inspecting both views might help detecting patterns in a data set. 
Additionally, 
we are using a special spacing and shading normally used to `highlight' %'
the second variable in the first (as will be discussed in Section \ref{sec:spacing}): here,
the intention of the spacing 
is to emphasize the conditional distributions of the second variable,
given the first one, and the shading helps identifying the factor levels in
the second variable.

<<pairs>>=
pairs(hec, highlighting = 2, diag_panel = pairs_diagonal_mosaic, 
      diag_panel_args = list(fill = grey.colors))
@

% auch möglich mit Titanic-Daten:
% <<pairs>>=
% pairs(tit, highlighting = 2, diag_panel = pairs_diagonal_mosaic, 
%       diag_panel_args = list(fill = grey.colors, alternate_labels = FALSE))
% @

%\setkeys{Gin}{width=\textwidth}
\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<pairsfig,fig=TRUE,echo=FALSE,width=8,height=8>>=
<<pairs>>
@
\caption{Pairs plot for the \data{HairEyeColor} data.}
\label{fig:pairs}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\begin{center}
\code{cell:\emph{Variable1}=\emph{Level1},\emph{Variable2}=\emph{Level2}} \dots
\end{center}

@@

>>>>

<<subsetting,results=verbatim>>=
(tit <- structable(~ Sex + Class + Age, data = Titanic[,,2:1,]))
tit["Male",]
tit["Male", c("1st","2nd","3rd")]
@

@@

@@

<<cotabplot>>=
cotabplot(~ Class + Age | Sex, data = tit, split_vertical = TRUE)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=11>>=
<<cotabplot>>
@
\caption{Conditional table plot for the \data{Titanic} data,
  again visualizing the distribution of age and class, given
  gender, using separate mosaic displays like the `manual' plot 
  in Figure \ref{fig:parttable}.}
\label{fig:cotabplot}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}
%\noindent The \code{labeling\_args} argument modifies the labels'
%appearance: here, to be left-aligned and unclipped
%(see Section \ref{sec:labeling}).

\noindent Visualizing the strata separately `hides' the distribution of the
conditioning variable(s) which may or may not be appropriated or
sensible in a particular analysis step. If we wish to keep the
information on the marginal distribution(s), we can use one single
mosaic for the stratified plot since mosaic displays are 
``conditional plots'' by definition. We just need to make sure that
conditioning variables are used first for splitting. 
Both the default and the formula interface of \codefun{mosaic} allow the
specification of conditioning variables (see Figure \ref{fig:conditioning}):

<<Conditioning1>>=
mosaic(tit, conditional = "Sex", split_vertical = c(TRUE, TRUE, FALSE), 
       keep_aspect_ratio = FALSE)
@ 
<<Conditioning2>>=
mosaic(~ Class + Age | Sex, data = tit, split_vertical = c(TRUE, TRUE, FALSE), 
       keep_aspect_ratio = FALSE)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<Conditioningfig,fig=TRUE,echo=FALSE,width=12>>=
<<Conditioning2>>
@
\caption{Mosaic plot again visualizing the 
distribution of class and age, given gender, this time using a
single mosaic plot. In contrast to Figure \ref{fig:parttable}, this
plot also visualizes the marginal distribution of `Sex'.}
\label{fig:conditioning}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7}

\noindent The effect of using this is that
conditioning variables are permuted ahead of the the conditioned variables
in the table, and for the formula interface, 
that \codefun{spacing\_conditional} is used as default to better
distinguish conditioning from conditioned dimensions.
This spacing uses equal space between tiles of conditioned variables,
and increasing space between tiles of conditioning variables. In
addition, we release the fixed aspect ratio to get less distorted margins.

Another high-level function for visualizing conditional independence
models are the \codefun{pairs} methods for table and structable objects.
In contrast to \codefun{cotabplot} which conditions on variables,
the \codefun{pairs} methods create pairwise views of the table. 
The function produces, by default, a plot matrix 
having strucplot displays in the off-diagonal panels, and
the variable names (optionally, with univariate statistics) in the diagonal cells. 
Figure \ref{fig:pairs} shows a pairs display for the
\data{Titanic} data with univariate mosaics in the diagonal,
and mosaic plots visualizing mutual independence in the upper and lower triangles.
Due to the inherent asymmetry of mosaic displays, the corresponding plots in the
upper and lower triangle differ depending on which variable is used first
for splitting---inspecting both views might help detecting patterns in a data set. 
Additionally, 
we are using a special spacing and shading normally used to `highlight' %'
the second variable in the first (as will be discussed in Section \ref{sec:spacing}): here,
the intention of the spacing 
is to emphasize the conditional distributions of the second variable,
given the first one, and the shading helps identifying the factor levels in
the second variable.

<<pairs>>=
pairs(tit, highlighting = 2, diag_panel = pairs_diagonal_mosaic, 
      diag_panel_args = list(fill = grey.colors, alternate_labels = FALSE))
@

%\setkeys{Gin}{width=\textwidth}
\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<pairsfig,fig=TRUE,echo=FALSE,width=8,height=8>>=
<<pairs>>
@
\caption{Pairs plot for the \data{Titanic} data.}
\label{fig:pairs}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\begin{center}
\code{[Optional prefix]cell:\emph{Variable1}=\emph{Level1},\emph{Variable2}=\emph{Level2}} \dots
\end{center}

@@

\subsection{Performance Issues}
\label{sec:performance}

As stated above, the implementation of strucplot displays is based on
creating and nesting \pkg{grid} viewports. The main time-consuming steps performed by
the core functions are the follwing:

\begin{enumerate}
\item recursively, create a tree of nested viewports
\item push the whole tree, i.e., add it to the system tree maintained
  by the \pkg{grid} engine
\item one by one, seek all cells of the strucplot display and add
  plot-specific content (rectangles for mosaics, bars for association
  plots, etc.)
\end{enumerate}

\noindent All these operations scale linearly with the amount of
created viewports. For a $d$-dimensional table with $k_i$
levels, $i=1 \dots d$, the total number of needed viewports
$T_d$ can roughly be estimated as:

\begin{equation}
  \label{eq:numbervp}
  T_d = \sum_{i=1}_d \prod_{j \le i} k_j
\end{equation}

\noindent Assuming an equal number for all dimensions, we obtain:

\begin{equation}
  \label{eq:equalvp}
  T_d = \sum_{i=1}_d k^i = \frac{k(k^d-1)}{k-1}
\end{equation}

\noindent and so the time complexity for drawing a strucplot display is of order $O(k^d)$.
